\large \texttt{insectDisease}: programmatic access to the \textit{Ecological Database of the World's Insect Pathogens}

### 2022-08-25 manuscript/manuscript.tex
fixed the Rbuildignore and the gitignore
updates to manuscript before send off, CRAN badge added to README

%%% Target journal
% Ecography (software note) -- maybe we shouldn't re-submit and pay $3000 to publish? 
% Journal of Open Source Software 
% Ecology (data paper)
%

%%% Potential reviewers
% Lewis Bartlett  lewis.Bartlett@uga.edu UGA
% Philipp Boersch  pboesu@gmail.com   British Trust for Ornithology
% Ben Longdon b.longdon2@exeter.ac.uk Uni of Exeter
% Susan Braxton braxton@illinois.edu U Illinois (could be COI with Onstad though)
% 
%

%%% Editors
% Thiago Rangel
% 
%

@@

\clearpage
\setstretch{1.35}
\linenumbers

@@

\clearpage
\setstretch{1.9}
\section*{Introduction}

@@

\subsection*{Solution statement}
To preserve these data in a format that is well-documented, openly accessible, versioned, and flexible for continued development, we created the \texttt{insectDisease} R package. In doing so, we implicitly adhere to the FAIR (Findable, Accessible, Interoperable, Resuable) guidelines for managing data \citep{wilkinson2016}. By hosting the data openly on GitHub, and versioning releases of the data with a permanent identifier (DOI), we ensure the longevity and versioned curation of this data resource. Finally, the incorporation of taxonomic data through \texttt{taxize} \citep{chamberlain2013} ensures that host and pathogen taxonomic names are updated periodically to accommodate for dynamic data or changing taxonomies.

@@

\paragraph*{}
Maintaining the data dynamically as described above allows users to access the data programmatically or as versioned flatfiles (i.e., \texttt{.csv} files). However, for users who do not wish to download the entire data resource, and simply want to quickly query a static version of the database, there is also a standalone web user interface (\texttt{https://edwip.ecology.uga.edu/}) that allows users to easily subset and explore the data. The web interface serves arguably the most important subset of the overall data (data files \texttt{nematode}, \texttt{viruses},\texttt{nvpassoc}, \texttt{negative}, and \texttt{hosts}). This interface allows users to quickly query based on host or parasite taxonomy as a dropdown list. This is perhaps more useful as a teaching tool or for initial exploration of the data, while the programmatic interface and dynamic data may be more useful for more rigorous analysis. This version of the EDWIP data will also only be deployed with a single static copy of the data, such that users wanting to benefit from versioned and dynamic data will need to access the data through the GitHub repository. Future efforts to integrate the web interface and the existing dynamic data structure will be explored, but this is not currently integrated.

@@

\subsection*{Data accessibility}
The \texttt{insectDisease} R package is currently available on GitHub \\ (github.com/viralemergence/insectDisease), with `.csv' files in the csv directory for long-term data stability. GitHub releases of the data ensure versioning is maintained and all versions are accessible. At the time of this writing, the current version is 1.2.0 (available at \\ \texttt{https://github.com/viralemergence/insectDisease/releases/tag/1.2.0}). Releases are given a DOI through integration with Zenodo (\texttt{https://doi.org/10.5281/zenodo.5821896}).

@@

\begin{table}[ht]
\centering
\caption{Files associated with the EDWIP data resource. Metadata is stored in $R$ package documentation, allowing the data and metadata to be intrinsically linked. For instance, users can use the help functionality from within $R$ to see more information on data columns and unit (e.g., \texttt{?nematode}). }
\label{tab:data}
\begin{tabular}{cccp{8cm}}
  \hline
  filename & rows & columns & description \\
  \hline
   \texttt{assocref} & 11005  & 16 & references for some host-pathogen associations  \\
   \texttt{citation} & 1966  & 7  & references but no host-pathogen association information  \\
   \texttt{hosts} & 4392  & 21  & insect host trait data\\   
   \texttt{hostTaxonomy} & 4489  & 7  & host taxonomic data updated with the \texttt{getNCBI()} function \\
   \texttt{negative} & 529  & 21  & information on negative host-pathogen associations  \\
   \texttt{nemaref} & 338  & 5  & references from nematode pathogens  \\
   \texttt{nematode} & 234  & 24  & host-nematode interaction data \\   
   \texttt{new\_asso} & 19  & 25  & likely a training document (perhaps do not use)  \\ 
   \texttt{noassref} & 569  & 16  &  references for some host-pathogen associations  \\
   \texttt{nvpassoc} & 7164  & 23  & non-viral pathogen infection data  \\   
   \texttt{pathogen} & 2041  & 9  &  pathogen trait data  \\   
   \texttt{pathTaxonomy} & 2282  & 7  & pathogen taxonomic data updated with the \texttt{getNCBI()} function  \\
   \texttt{viraref} & 2124  & 16  & references from viral infections  \\   
   \texttt{viruses} & 1659  & 25  & host-viral interaction data  \\      
  \hline
\end{tabular}
\end{table}

>>>>

@@

\clearpage
\setstretch{1.25}
\linenumbers

\noindent \texttt{insectDisease}: programmatic access to the \textit{Ecological Database of the World's Insect Pathogens}

@@

\setstretch{1.9}
\section*{Introduction}

@@

\subsection*{Solution statement}
To preserve these data in a format that is well-documented, openly accessible, versioned, and flexible for continued development, we created the \texttt{insectDisease} R package. In doing so, we implicitly adhere to the FAIR (Findable, Accessible, Interoperable, Resuable) guidelines for managing data \citep{wilkinson2016}. By hosting the data openly on GitHub (\texttt{https://github.com/viralemergence/insectDisease/}) -- versioning releases of the data with a permanent identifier (DOI) on Zenodo -- and on CRAN (\texttt{https://cran.r-project.org/web/packages/insectDisease/index.html}), we ensure the longevity and versioned curation of this data resource. Finally, the incorporation of taxonomic data through \texttt{taxize} \citep{chamberlain2013} ensures that host and pathogen taxonomic names are updated periodically to accommodate for dynamic data or changing taxonomies.

@@

\paragraph*{}
Maintaining the data dynamically as described above allows users to access the data programmatically or as versioned flatfiles (i.e., \texttt{.csv} files). However, for users who do not wish to download the entire data resource, and simply want to quickly query a static version of the database, there is also a standalone web user interface (\texttt{https://edwip.ecology.uga.edu/}) that allows users to easily subset and explore the data. The web interface serves arguably the most important subset of the overall data (data files \texttt{nematode}, \ \texttt{viruses}, \ \texttt{nvpassoc}, \ \texttt{negative}, and \ \texttt{hosts}). This interface allows users to quickly query based on host or parasite taxonomy as a dropdown list. This is perhaps more useful as a teaching tool or for initial exploration of the data, while the programmatic interface and dynamic data may be more useful for more rigorous analysis. This version of the EDWIP data will also only be deployed with a single static copy of the data, such that users wanting to benefit from versioned and dynamic data will need to access the data through the GitHub repository. Future efforts to integrate the web interface and the existing dynamic data structure will be explored, but this is not currently integrated.

@@

\subsection*{Data accessibility}
The \texttt{insectDisease} R package is currently available on GitHub \\ (github.com/viralemergence/insectDisease), with `.csv' files in the csv directory for long-term data stability. GitHub releases of the data ensure versioning is maintained and all versions are accessible. At the time of this writing, the current version is 1.2.1 (available at \\ \texttt{https://github.com/viralemergence/insectDisease/releases/tag/1.2.1}). \\
Releases are given a DOI through integration with Zenodo (available at \\ \texttt{https://doi.org/10.5281/zenodo.5821896}). The package (release version 1.2.1) is also available on CRAN \\ \texttt{https://cran.r-project.org/web/packages/insectDisease/index.html}.

@@

\begin{table}[ht]
\centering
\caption{Files associated with the EDWIP data resource. Metadata is stored in $R$ package documentation, allowing the data and metadata to be intrinsically linked. For instance, users can use the help functionality from within $R$ to see more information on data columns and unit (e.g., \texttt{?nematode}). }
\label{tab:data}
\begin{tabular}{cccp{8cm}}
  \hline
  filename & rows & columns & description \\
  \hline
   \texttt{assocref} & 11005  & 16 & references for some host-pathogen associations  \\
   \texttt{citation} & 1966  & 7  & references but no host-pathogen association information  \\
   \texttt{hosts} & 4392  & 21  & insect host trait data\\   
   \texttt{hostTaxonomy} & 4489  & 7  & host taxonomic data updated with the \texttt{getNCBI()} function \\
   \texttt{negative} & 529  & 21  & information on negative host-pathogen associations  \\
   \texttt{nemaref} & 338  & 16  & references from nematode pathogens  \\
   \texttt{nematode} & 234  & 24  & host-nematode interaction data \\   
   \texttt{new\_asso} & 19  & 25  & likely a training document (perhaps do not use)  \\ 
   \texttt{noassref} & 569  & 16  &  references for some host-pathogen associations  \\
   \texttt{nvpassoc} & 7164  & 23  & non-viral pathogen infection data  \\   
   \texttt{pathogen} & 2041  & 9  &  pathogen trait data  \\   
   \texttt{pathTaxonomy} & 2282  & 7  & pathogen taxonomic data updated with the \texttt{getNCBI()} function  \\
   \texttt{viraref} & 2124  & 16  & references from viral infections  \\   
   \texttt{viruses} & 1659  & 26  & host-viral interaction data  \\      
  \hline
\end{tabular}
\end{table}
